---
title: "Day 3: Single-layered neural networks (Exercises with answers)"
author: "Emilio Lehoucq"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
params:
  notes: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.height = 4.35, fig.width = 4.75, message = FALSE, warning = FALSE)
```

# Question

Fit a single-layered neural network through 2 repeats of 3-fold cross validation to predict `order` in the [clickstream data for online shopping Data Set](https://archive.ics.uci.edu/ml/datasets/clickstream+data+for+online+shopping). Which model fits the data best? Use a validation set approach. Also, visualize the marginal relation between `day` and the response with an ALE plot.

# Answer

## Packages

```{r}
library(tidyverse)
library(janitor) # clean_names
library(robustHD) # standardize
library(nnet) # neural networks
library(caret) # cross validation
library(ALEPlot) # ALE plot
```

## Data

```{r}
set.seed(1) 

factors <- c("country", "page_1_main_category", "colour", 
             "location", "model_photography", "price_2", "page")

eshop <- read_delim(here::here("data/e-shop clothing 2008.csv"), delim = ";") %>% 
  sample_n(1500) %>% 
  clean_names() %>% 
  select(-c(`session_id`, `page_2_clothing_model`, year)) %>% 
  mutate_each_(funs(factor(.)), factors) %>% 
  mutate_if(is.numeric, funs(standardize(.)))
```

```{r}
set.seed(1)

eshop_train <- eshop %>% sample_frac(0.7)
eshop_test <- eshop %>% setdiff(eshop_train)
```

## Single-layered neural network

```{r, echo=TRUE}
set.seed(1)

folds <- createFolds(eshop_train$order, k = 3, list = FALSE)

training <- eshop_train[ folds,]

testing  <- eshop_train[-folds,]

train_control <- trainControl(method = "repeatedcv",
                           number = 3,
                           repeats = 2)

nn <- train(order ~ ., data = training, 
            method = "nnet", 
            trControl = train_control,
            verbose = FALSE,
            trace = FALSE,
            tuneGrid = expand.grid(decay = c(0.1, 0.5, 1, 1.5), size = c(1, 5, 10, 15, 19)))
```

```{r}
trellis.par.set(caretTheme())
ggplot(nn)  
```

## ALE plot

```{r, echo=TRUE}
nn_best <- nnet(order~.,eshop_train, linout=T, size=15, decay=0.1, maxit=1000, trace=F)

yhatf <- function(X.model, newdata) as.numeric(predict(X.model, newdata))

invisible(ALEPlot(as.data.frame(eshop_train), nn_best, pred.fun=yhatf, J=2, K=50, NA.plot = TRUE))
rug(pull(eshop_train))
```

## Model validation

```{r}
# helper function
#' @name r2
#' @param model nnet object
#' @param test test data
#' @param outcome string with the response
#' @returns test R2
r2<- function(model, test, outcome){
  
  preds <- predict(model, test)
  
  sse <- sum((test[[outcome]] - preds)^2)
  sst <- sum((test[[outcome]] - mean(test[[outcome]]))^2)
  r2 <- 1 - sse / sst
  
  return(r2)
  
}
```

```{r}
r2(nn_best, eshop_test, "order")
```

| Model            | Test R2 |
|------------------|---------|
| Kitchen sink OLS | 0.08    |
| Ridge            | 0.08    |
| Lasso            | 0.09    |
| Random forests   | 0.03    |
| Boosted trees    | -0.06   |
| Neural network   | -0.97   |
