---
title: "Day 3: Single-layered neural networks (Exercises with answers)"
author: "Emilio Lehoucq"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
params:
  notes: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.height = 4.35, fig.width = 4.75, message = FALSE, warning = FALSE)
```

# Question

Fit a single-layered neural network through 2 repeats of 3-fold cross validation to predict `order` in the [clickstream data for online shopping Data Set](https://archive.ics.uci.edu/ml/datasets/clickstream+data+for+online+shopping). Which model fits the data best? Use a validation set approach. Also, visualize the marginal relation between `day` and the response with an ALE plot.

# Answer

## Packages

```{r}
library(tidyverse)
library(janitor)
library(robustHD)
library(nnet)
library(caret)
library(ALEPlot)
```

## Data

```{r}
set.seed(1) 

factors <- c("country", "page_1_main_category", "colour", 
             "location", "model_photography", "price_2", "page")

eshop <- read_delim(here::here("data/e-shop clothing 2008.csv"), delim = ";") %>% 
  sample_n(1500) %>% 
  clean_names() %>% 
  select(-c(`session_id`, `page_2_clothing_model`, year)) %>% 
  mutate_each_(funs(factor(.)), factors) %>% 
  mutate(log_order = log(order)) %>%
  select(-order) %>% 
  mutate_if(is.numeric, funs(standardize(.)))
```

```{r}
set.seed(1)

eshop_train <- eshop %>% sample_frac(0.7)
eshop_test <- eshop %>% setdiff(eshop_train)
```

## Single-layered neural network

```{r, echo=TRUE}
set.seed(1)

train_control <- trainControl(method = "repeatedcv", number = 3, repeats = 2)

nn <- train(log_order ~ ., data = eshop_train, method = "nnet", trControl = train_control, verbose = FALSE, trace = FALSE,
            tuneGrid = expand.grid(decay = c(1, 10, 100, 1000), size = c(1, 2, 3, 4, 5)))

print(nn)
```

```{r}
trellis.par.set(caretTheme())
ggplot(nn, metric = "Rsquared")  
```

## ALE plot

```{r, echo=TRUE}
nn_best1 <- nnet(log_order~.,eshop_train, linout=T, size=1, decay=1, maxit=1000, trace=F)

yhatf <- function(X.model, newdata) as.numeric(predict(X.model, newdata))

invisible(ALEPlot(as.data.frame(eshop_train), nn_best1, pred.fun=yhatf, J=2, K=50, NA.plot = TRUE))
rug(pull(eshop_train))
```

```{r, echo=TRUE}
nn_best2 <- nnet(log_order~.,eshop_train, linout=T, size=3, decay=1, maxit=1000, trace=F)

invisible(ALEPlot(as.data.frame(eshop_train), nn_best2, pred.fun=yhatf, J=2, K=50, NA.plot = TRUE))
rug(pull(eshop_train))
```

## Model validation

```{r}
# helper function
#' @name r2
#' @param model nnet object
#' @param test test data
#' @param outcome string with the response
#' @returns test R2
r2<- function(model, test, outcome){
  
  preds <- predict(model, test)
  
  sse <- sum((test[[outcome]] - preds)^2)
  sst <- sum((test[[outcome]] - mean(test[[outcome]]))^2)
  r2 <- 1 - sse / sst
  
  return(r2)
  
}
```

```{r}
r2(nn_best1, eshop_test, "log_order")
```

```{r}
r2(nn_best2, eshop_test, "log_order")
```

| Model            | Test R2 |
|------------------|---------|
| Kitchen sink OLS | 0.08    |
| Ridge            | 0.08    |
| Lasso            | 0.09    |
| Random forests   | 0.03    |
| Boosted trees    | 0.01    |
| Neural network   | 0.09    |
